---
title: "Conditional Independencies of a GRN"
output:
  html_document: default
  pdf_document: 
    fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```


```{r , message= FALSE}
library(dagitty)
library(lavaan)
library(sna)
library(DOT)
library(igraph)
library(pcalg)
library(SEMgraph)
```

# Simulation example 1

```{r}
g<- dagitty("dag {
    X -> R -> S -> T <- U <- V -> Y
    T -> P
}")
g<-graphLayout(g)
plot(g)
```

### For each non-adjacent pair of variables, all minimal sets that we can condition on to render that pair independent

```{r}
impliedConditionalIndependencies(g, type = "missing.edge", max.results = Inf)
```

### Independence of X and Y  given {R,V}

```{r}
paths(g, "X", "Y", c("R","V") ) 
```
X and Y are independent because the path between them is closed given {R,V}

### All pairs of variables in the graph that are independent conditional on the set Z={R,V}

```{r}
pairs <- combn( names(g), 2 )
apply( pairs, 2, function(x){
  p <- paths(g, x[1], x[2], c("R","V") )
  if( !p$open ){
    message( x[1]," and ",x[2]," are independent given {R,V}" )
  } else {
    message( x[1]," and ",x[2]," are possibly dependent given {R,V}" )
  }
} )
```

### For each pair of non-adjacent nodes in the graph, determine whether they are independent conditional on all other variables

```{r}
apply( pairs, 2, function(x){
  all.other.variables <- setdiff( names(g), x )
  if( dseparated(g, x[1], x[2], all.other.variables ) ){
    message( x[1]," and ",x[2]," are independent given ", 
             paste( all.other.variables, collapse=",") )
  }
} )
```

### Regression coefficients vs structural causal models

Fitting models with fixed effects (no latent variables)
Y=a+bX+cR+dS+eT+fP

```{r}
predictors <- c("X","R","S","T","P")
```

Which of the five variables X,R,S,T,P are d-separated from Y given the same five variables?
(i.e., P(X,Y|Z)=P(X|Z)P(Y|Z)?)

```{r}
dseparated( g, "Y", list(), predictors)
intersect(predictors, dseparated( g, "Y", list(), predictors)) #Result: X,R,P
```

Simulate data

```{r}
d <- simulateSEM( g, .7, .7, N=10000 )
head(d)
fit.lm <- lm( Y ~ X + R + S + T + P, data=d )
confint(fit.lm) #Result: X,R,P coeficients are zero. The conditional independence is verified
```


# Simulation example 2

### The Causal Interpretation of Structural Coefficients

```{r}
N <- 10000 # sample size
Ux <- rnorm( N )
Uy <- rnorm( N )
Uz <- rnorm( N )
X <- Ux
Y <- 1/3*X + Uy
Z <- 1/16*Y + Uz
d <- data.frame(X=X,Y=Y,Z=Z)

g <- dagitty("dag {
    Ux -> X -> Y -> Z <- Uz
    Uy -> Y
}")
coordinates(g) <- list(
  x=c(Ux=1,Uy=2,Uz=3,X=1,Y=2,Z=3),
  y=c(Ux=1,Uy=1,Uz=1,X=0,Y=0,Z=0) )
plot(g)
```

### For each non-adjacent pair of variables, all minimal sets that we can condition on to render that pair independent

```{r}
impliedConditionalIndependencies(g, type = "missing.edge", max.results = Inf)
```

### For each pair of non-adjacent nodes in the graph, determine whether they are independent conditional on all other variables

```{r}
pairs <- combn( names(g), 2 )
apply( pairs, 2, function(x){
  all.other.variables <- setdiff( names(g), x )
  if( dseparated(g, x[1], x[2], all.other.variables ) ){
    message( x[1]," and ",x[2]," are independent given ", 
             paste( all.other.variables, collapse=",") )
  }
} )
```

# E. coli GRN

```{r,fig.width=9, fig.height=11}
dot <- read.dot("~/Missing data/SERGIO/GNW_sampled_GRNs/Ecoli_100_net1.dot")
g.graph <- graph.adjacency(dot)
plot(g.graph,vertex.size=5, edge.arrow.size=.2)
g <- graph2dagitty(g.graph)
g <- graphLayout(g)
plot(g)
```

### Conditional independencies

For each non-adjacent pair of variables, all minimal sets that we can condition on to render that pair independent

```{r}
impliedConditionalIndependencies(g, type = "missing.edge", max.results = Inf)
```


Draw paths from icd to fadI

```{r}
paths(g, "icd", "fadI" )$paths
```

Draw directed paths from icd to fadI

```{r}
paths(g, "potI", "nemA", directed=TRUE)$paths
```

