---
title: "Case Study 1 : The multicause model"
author: ""
date: "1/20/2021"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(rstan)
library(ggplot2)
library(dplyr)
```

```{r}
D = 200
startSeed = 1
K = 10
ndp = c(15,20,60,100)
x1 = rep(0,D)
x2 = rep(0,D)
x3 = rep(0,D)
x4 = rep(0,D)
x5 = rep(0,D)
num_generated_samples = D
```


# Create observational data

In this part we generate K observational data where each data has a fixed seed.

```{r}
U1mean = -8
U1Var = 10
U2mean = 6
U2Var = 8
U3mean = -5
U3Var = 7

theta_zero_X1 = 2
theta_U1X1 = 0.23
theta_U2X1 = 0.33
theta_U3X1 = 0.12

theta_zero_X2 = 1.59
theta_U1X2 = 0.23
theta_U2X2 = 0.34
theta_U3X2 = 0.15

theta_zero_X3 = 1.52
theta_U1X3 = 0.21
theta_U2X3 = 0.32
theta_U3X3 = 0.13

theta_zero_X4 = 1.49
theta_U1X4 = 0.24
theta_U2X4 = 0.35
theta_U3X4 = 0.16

theta_zero_X5 = 1.38
theta_U1X5 = 0.28
theta_U2X5 = 0.38
theta_U3X5 = 0.10

theta_zero_M = -1.4
theta_X1M = 0.36
theta_X2M = 0.12
theta_X3M = 0.10
theta_X4M = 0.11
theta_X5M = 0.26

theta_zero_Y = -1.1
theta_MY = 0.26
theta_U1Y = 0.22
theta_U2Y = 0.15
theta_U3Y = 0.11
```


```{r}
obs_data_list <- list()
data_count = 1

for (seed in startSeed:50) {
    set.seed(seed)
    U1 = rnorm(D, U1mean, U1Var)
    U2 = rnorm(D, U2mean, U2Var)
    U3 = rnorm(D, U3mean, U3Var)
    
    p1_train = rep(0,D)
    p2_train = rep(0,D)
    p3_train = rep(0,D)
    p4_train = rep(0,D)
    p5_train = rep(0,D)
    p1_train = 1/(1+exp(-theta_zero_X1 - theta_U1X1 * U1 - theta_U2X1 * U2 - theta_U3X1 * U3))
    p2_train = 1/(1+exp(-theta_zero_X2 - theta_U1X2 * U1 - theta_U2X2 * U2 - theta_U3X2 * U3))
    p3_train = 1/(1+exp(-theta_zero_X3 - theta_U1X3 * U1 - theta_U2X3 * U2 - theta_U3X3 * U3))
    p4_train = 1/(1+exp(-theta_zero_X4 - theta_U1X4 * U1 - theta_U2X4 * U2 - theta_U3X4 * U3))
    p5_train = 1/(1+exp(-theta_zero_X5 - theta_U1X5 * U1 - theta_U2X5 * U2 - theta_U3X5 * U3))
    X1 <- rbinom(n = D, size = 1, prob = p1_train)
    X2 <- rbinom(n = D, size = 1, prob = p2_train)
    X3 <- rbinom(n = D, size = 1, prob = p3_train)
    X4 <- rbinom(n = D, size = 1, prob = p4_train)
    X5 <- rbinom(n = D, size = 1, prob = p5_train)
    
    pm = rep(0, D)
    pm = 1/(1+exp(-theta_zero_M - theta_X1M * X1- theta_X2M * X2- theta_X3M * X3- theta_X4M * X4- theta_X5M * X5))
    M <- rbinom(n = D, size = 1, prob = pm)
    
    py = rep(0, D)
    py = 1/(1+exp(-theta_zero_Y - theta_MY * M - theta_U1Y * U1 - theta_U2Y * U2 - theta_U3Y * U3))
    Y <- rbinom(n = D, size = 1, prob = py)
    obs_data_list[[data_count]] <- data.frame("U1" = U1,"U2" = U2, "U3" = U3, "X1" = X1, "X2" = X2,"X3" = X3,"X4" = X4,"X5" = X5, "M" = M, "Y" = Y)
  data_count = data_count + 1
}
```

# Create interventional data

```{r}
intv_data_list <- list()
data_count = 1

for (seed in startSeed:50) {
    set.seed(seed)
    U1 = rnorm(D, U1mean , U1Var )
    U2 = rnorm(D,U2mean,U2Var)
    U3 = rnorm(D, U3mean, U3Var)
    
    X1 <- x1
    X2 <- x2
    X3 <- x3
    X4 <- x4
    X5 <- x5
    
    pm = rep(0, D)
    pm = 1/(1+exp(-theta_zero_M - theta_X1M * X1- theta_X2M * X2- theta_X3M * X3- theta_X4M * X4- theta_X5M * X5))
    M <- rbinom(n = D, size = 1, prob = pm)
    
    py = rep(0, D)
    py = 1/(1+exp(-theta_zero_Y - theta_MY * M - theta_U1Y * U1 - theta_U2Y * U2 - theta_U3Y * U3))
    Y <- rbinom(n = D, size = 1, prob = py)
    intv_data_list[[data_count]] <- data.frame("U1" = U1,"U2" = U2, "U3" = U3, "X1" = X1, "X2" = X2,"X3" = X3,"X4" = X4,"X5" = X5, "M" = M, "Y" = Y)
  data_count = data_count + 1
}
```

# True Stan model

```{r}
model_str <- "
    data {
        int D;
        int<lower=0,upper=1> X1[D];
        int<lower=0,upper=1> X2[D];
        int<lower=0,upper=1> X3[D];
        int<lower=0,upper=1> X4[D];
        int<lower=0,upper=1> X5[D];
        int<lower=0,upper=1> M[D];
        int<lower=0,upper=1> Y[D];
    }
  parameters {
       real<lower=0> u1_scale;
       real<lower=0> mu1;
       real<lower=0> u2_scale;
       real mu2;
       real<lower=0> u3_scale;
       real mu3;
       real theta_zero_X1; 
       real theta_U1X1;
       real theta_U2X1;
       real theta_U3X1;
       
       real theta_zero_X2; 
       real theta_U1X2;
       real theta_U2X2;
       real theta_U3X2;
       
       real theta_zero_X3; 
       real theta_U1X3;
       real theta_U2X3;
       real theta_U3X3;
       
       real theta_zero_X4; 
       real theta_U1X4;
       real theta_U2X4;
       real theta_U3X4;
       
       real theta_zero_X5; 
       real theta_U1X5;
       real theta_U2X5;
       real theta_U3X5;
       
       real theta_zero_M; 
       real theta_X1M;
       real theta_X2M; 
       real theta_X3M;
       real theta_X4M;
       real theta_X5M;
       real theta_zero_Y;
       real theta_MY;
       real theta_U1Y;
       real theta_U2Y;
       real theta_U3Y;
       vector[D] U1;
       vector[D] U2;
       vector[D] U3;
  }
    model {
        //prior over parameters
        u1_scale  ~ normal(0, 10); 
        u2_scale  ~ normal(0, 10); 
        u3_scale  ~ normal(0, 10); 
        mu1  ~ normal(0, 10); 
        mu2  ~ normal(0, 10); 
        mu3 ~ normal(0, 10);
        theta_zero_X1  ~ normal(0, 10); 
        theta_U1X1 ~ normal(0, 10); 
        theta_U2X1 ~ normal(0, 10); 
        theta_U3X1 ~ normal(0, 10);
        
        theta_zero_X2  ~ normal(0, 10); 
        theta_U1X2 ~ normal(0, 10); 
        theta_U2X2 ~ normal(0, 10); 
        theta_U3X2 ~ normal(0, 10); 
        
        theta_zero_X3  ~ normal(0, 10); 
        theta_U1X3 ~ normal(0, 10); 
        theta_U2X3 ~ normal(0, 10); 
        theta_U3X3 ~ normal(0, 10); 
        
        theta_zero_X4  ~ normal(0, 10); 
        theta_U1X4 ~ normal(0, 10); 
        theta_U2X4 ~ normal(0, 10); 
        theta_U3X4 ~ normal(0, 10); 
        
        theta_zero_X5  ~ normal(0, 10); 
        theta_U1X5 ~ normal(0, 10); 
        theta_U2X5 ~ normal(0, 10); 
        theta_U3X5 ~ normal(0, 10); 
        
        theta_zero_M ~ normal(0, 10); 
        theta_X1M ~ normal(0, 10); 
        theta_X2M ~ normal(0, 10); 
        theta_X3M ~ normal(0, 10); 
        theta_X4M ~ normal(0, 10); 
        theta_X5M ~ normal(0, 10); 
        
        theta_zero_Y ~ normal(0, 10); 
        theta_MY ~ normal(0, 10); 
        theta_U1Y ~ normal(0, 10); 
        theta_U2Y ~ normal(0, 10); 
        theta_U3Y ~ normal(0, 10);
        //likelihood
        U1 ~ normal(mu1, u1_scale); 
        U2 ~ normal(mu2, u2_scale);
        U3 ~ normal(mu3, u3_scale);
        X1 ~ bernoulli_logit(theta_zero_X1 + theta_U1X1 * U1 + theta_U2X1 * U2 + theta_U3X1 * U3); 
        X2 ~ bernoulli_logit(theta_zero_X2 + theta_U1X2 * U1 + theta_U2X2 * U2 + theta_U3X2 * U3); 
        X3 ~ bernoulli_logit(theta_zero_X3 + theta_U1X3 * U1 + theta_U2X3 * U2 + theta_U3X3 * U3); 
        X4 ~ bernoulli_logit(theta_zero_X4 + theta_U1X4 * U1 + theta_U2X4 * U2 + theta_U3X4 * U3); 
        X5 ~ bernoulli_logit(theta_zero_X5 + theta_U1X5 * U1 + theta_U2X5 * U2 + theta_U3X5 * U3); 
        for(i in 1:D) {
            M[i] ~ bernoulli_logit(theta_zero_M + theta_X1M * X1[i] + theta_X2M * X2[i] + theta_X3M * X3[i] + theta_X4M * X4[i] + theta_X5M * X5[i]);  
            Y[i] ~ bernoulli_logit(theta_zero_Y + theta_MY * M[i] + theta_U1Y * U1[i] + theta_U2Y * U2[i] + theta_U3Y * U3[i]); 
        }
    }
"
```


```{r, message=FALSE, warning=FALSE}
mod <- rstan::stan_model(model_code = model_str)
mod
```

# Mutilated model

```{r}
mutilated_model <- function(mu1, mu2,mu3,
                            u1_scale, u2_scale, u3_scale,
                            theta_zero_X1, theta_U1X1, theta_U2X1,
                            theta_zero_X2,theta_U1X2,theta_U2X2,
                            theta_zero_X3,theta_U1X3,theta_U2X3,
                            theta_zero_X4,theta_U1X4,theta_U2X4,
                            theta_zero_X5,theta_U1X5,theta_U2X5,
                            theta_zero_M,theta_X1M,theta_X2M,theta_X3M,theta_X4M,theta_X5M,
                            theta_zero_Y, theta_MY,
                            theta_U1Y, theta_U2Y,theta_U3Y,
                            x1, x2, x3, x4, x5,
                            seed, num_generated_samples) {
    set.seed(seed)
    U1 = rnorm(num_generated_samples, mu1, u1_scale)
    U2 = rnorm(num_generated_samples, mu2, u2_scale)
    U3 = rnorm(num_generated_samples, mu3, u3_scale)
    X1 <- x1
    X2 <- x2
    X3 <- x3
    X4 <- x4
    X5 <- x5
    
    pm = rep(0, D)
    pm = 1/(1+exp(-theta_zero_M - theta_X1M * X1 - theta_X2M * X2 - theta_X3M * X3 - theta_X4M * X4 - theta_X5M * X5))
    M <- rbinom(n = D, size = 1, prob = pm)

    py = rep(0, D)
    py = 1/(1+exp(-theta_zero_Y - theta_MY * M - theta_U1Y * U1 - theta_U2Y * U2 - theta_U3Y * U3))
    Y <- rbinom(n = D, size = 1, prob = py)
  return(Y)
}
```

```{r}
estimated_y_hmc_list <- list()
hmc_fit_list <- list()
means_hmc <- c()
means_hmc_list <- list()
start_time <- Sys.time()
for (num_data_points in ndp) {
  print("num_data_points is:")
  print(num_data_points)
  for (data_count in 1:K) {
    print("data_count is:")
    print(data_count)
    data_list <- list(D=num_data_points,
                      X1 = obs_data_list[[data_count]]$X1[1:num_data_points],
                      X2 = obs_data_list[[data_count]]$X2[1:num_data_points],
                      X3 = obs_data_list[[data_count]]$X3[1:num_data_points],
                      X4 = obs_data_list[[data_count]]$X4[1:num_data_points],
                      X5 = obs_data_list[[data_count]]$X5[1:num_data_points],
                      M = obs_data_list[[data_count]]$M[1:num_data_points],
                      Y = obs_data_list[[data_count]]$Y[1:num_data_points])
    #If you get this error (Cluster setup failed. 2 of 2 workers failed to connect.) run these lines:
    # if (Sys.getenv("RSTUDIO") == "1" && !nzchar(Sys.getenv("RSTUDIO_TERM")) &&
    #     Sys.info()["sysname"] == "Darwin") {
    #   parallel:::setDefaultClusterOptions(setup_strategy = "sequential")
    # }
    hmc_fit_list[[data_count]] <- rstan::sampling(mod, data=data_list, chains = 2, iter = 3000, warmup = 1500, seed = 1, control = list(max_treedepth = 15)) #seed 1 gives nice results
    samples_hmc <- rstan::extract(hmc_fit_list[[data_count]], c("mu1","mu2", "mu3", "u1_scale","u2_scale","u3_scale",
                                                                "theta_zero_X1", "theta_U1X1", "theta_U2X1",
                                                                "theta_zero_X2", "theta_U1X2", "theta_U2X2",
                                                                "theta_zero_X3", "theta_U1X3", "theta_U2X3",
                                                                "theta_zero_X4", "theta_U1X4", "theta_U2X4",
                                                                "theta_zero_X5", "theta_U1X5", "theta_U2X5",
                                                                "theta_zero_M","theta_X1M", "theta_X2M", "theta_X3M","theta_X4M","theta_X5M",
                                                                "theta_zero_Y", "theta_MY","theta_U1Y","theta_U2Y", "theta_U3Y"))
    estimated_y_hmc_list[[data_count]] <-  mutilated_model(mu1 = mean(samples_hmc$mu1),
                                            mu2 = mean(samples_hmc$mu2),
                                            mu3 = mean(samples_hmc$mu3),
                                            u1_scale = mean(samples_hmc$u1_scale),
                                            u2_scale = mean(samples_hmc$u2_scale),
                                            u3_scale = mean(samples_hmc$u3_scale),
                                            theta_zero_M = mean(samples_hmc$theta_zero_M),
                                            theta_X1M = mean(samples_hmc$theta_X1M),
                                            theta_X2M = mean(samples_hmc$theta_X2M),
                                            theta_X3M = mean(samples_hmc$theta_X3M),
                                            theta_X4M = mean(samples_hmc$theta_X4M),
                                            theta_X5M = mean(samples_hmc$theta_X5M),
                                            theta_zero_Y = mean(samples_hmc$theta_zero_Y),
                                            theta_MY = mean(samples_hmc$theta_MY),
                                            theta_U1Y = mean(samples_hmc$theta_U1Y),
                                            theta_U2Y = mean(samples_hmc$theta_U2Y),
                                            theta_U3Y = mean(samples_hmc$theta_U3Y),
                                            x1 = x1[1:num_generated_samples],
                                            x2 = x2[1:num_generated_samples],
                                            x3 = x3[1:num_generated_samples],
                                            x4 = x4[1:num_generated_samples],
                                            x5 = x5[1:num_generated_samples],
                                            seed = seed,
                                            num_generated_samples = num_generated_samples
    )
    means_hmc <- c(means_hmc,abs(mean(estimated_y_hmc_list[[data_count]]) - mean(intv_data_list[[data_count]]$Y)))
    
  }
  means_hmc_list[[paste0("num_data_points_",num_data_points)]] <- means_hmc
  print("means_hmc_list is:")
  print(means_hmc_list)
  means_hmc <- c()
}
end_time <- Sys.time()
end_time - start_time
#saveRDS(means_hmc_list, file = "data/Case1/output/means_hmc_list_Case1.RData")
```


# Misspecified model

```{r}
model_str_misspecify <- "
    data {
        int D;
        int<lower=0,upper=1> X1[D];
        int<lower=0,upper=1> X2[D];
        int<lower=0,upper=1> X3[D];
        int<lower=0,upper=1> X4[D];
        int<lower=0,upper=1> X5[D];
        int<lower=0,upper=1> M[D];
        int<lower=0,upper=1> Y[D];
    }
  parameters {
       real<lower=0> u1_scale;
       real<lower=0> mu1;
       
       real theta_zero_X1; 
       real theta_U1X1;
       
       real theta_zero_X2; 
       real theta_U1X2;
       
       real theta_zero_X3; 
       real theta_U1X3;
       
       real theta_zero_X4; 
       real theta_U1X4;
       
       real theta_zero_X5; 
       real theta_U1X5;
       
       real theta_zero_M; 
       real theta_X1M;
       real theta_X2M; 
       real theta_X3M; 
       real theta_X4M; 
       real theta_X5M; 
       
       real theta_zero_Y;
       real theta_MY;
       
       real theta_U1Y;
       vector[D] U1;
       
    }

    model {
        //prior over parameters
        u1_scale  ~ normal(0, 10); 
        mu1  ~ normal(0, 10); 
        
        theta_zero_X1  ~ normal(0, 10); 
        theta_U1X1 ~ normal(0, 10); 
        
        theta_zero_X2  ~ normal(0, 10); 
        theta_U1X2 ~ normal(0, 10); 
        
        theta_zero_X3  ~ normal(0, 10); 
        theta_U1X3 ~ normal(0, 10); 
        
        theta_zero_X4  ~ normal(0, 10); 
        theta_U1X4 ~ normal(0, 10); 
        
        theta_zero_X5  ~ normal(0, 10); 
        theta_U1X5 ~ normal(0, 10); 
        
        theta_zero_M ~ normal(0, 10); 
        theta_X1M ~ normal(0, 10); 
        theta_X2M ~ normal(0, 10); 
        theta_X3M ~ normal(0, 10);
        theta_X4M ~ normal(0, 10);
        theta_X5M ~ normal(0, 10);
        
        theta_zero_Y ~ normal(0, 10); 
        theta_MY ~ normal(0, 10); 
        theta_U1Y ~ normal(0, 10); 
        //likelihood
        U1 ~ normal(mu1, u1_scale); 
        X1 ~ bernoulli_logit(theta_zero_X1 + theta_U1X1 * U1); 
        X2 ~ bernoulli_logit(theta_zero_X2 + theta_U1X2 * U1); 
        X3 ~ bernoulli_logit(theta_zero_X2 + theta_U1X2 * U1); 
        X4 ~ bernoulli_logit(theta_zero_X2 + theta_U1X2 * U1); 
        X5 ~ bernoulli_logit(theta_zero_X2 + theta_U1X2 * U1); 
        for(i in 1:D) {
            M[i] ~ bernoulli_logit(theta_zero_M + theta_X1M * X1[i] + theta_X2M * X2[i] + theta_X3M * X3[i] + theta_X4M * X4[i] + theta_X5M * X5[i]);  
            Y[i] ~ bernoulli_logit(theta_zero_Y + theta_MY * M[i] + theta_U1Y * U1[i]); 
        }
    }
"
```


```{r, message=FALSE, warning=FALSE}
mod_misspecify <- rstan::stan_model(model_code = model_str_misspecify)
```

# Mutilated model

```{r}
mutilated_model_misspecify <- function(mu1, u1_scale,
                                       theta_zero_X1, theta_U1X1, theta_U2X1,
                                       theta_zero_X2,theta_U1X2,theta_U2X2,
                                       theta_zero_X3,theta_U1X3,theta_U2X3,
                                       theta_zero_X4,theta_U1X4,theta_U2X4,
                                       theta_zero_X5,theta_U1X5,theta_U2X5,
                                       theta_zero_M,theta_X1M,theta_X2M,theta_X3M,theta_X4M,theta_X5M,
                                       theta_zero_Y, theta_MY,
                                       theta_U1Y, theta_U2Y,theta_U3Y,
                                       x1, x2, x3, x4, x5,
                                       seed, num_generated_samples) {
  set.seed(seed)
  U1 = rnorm(num_generated_samples, mu1, u1_scale)
  
  X1 <- x1
  X2 <- x2
  X3 <- x3
  X4 <- x4
  X5 <- x5
  
  pm = rep(0, D)
  pm = 1/(1+exp(-theta_zero_M - theta_X1M * X1 - theta_X2M * X2 - theta_X3M * X3 - theta_X4M * X4 - theta_X5M * X5))
  M <- rbinom(n = D, size = 1, prob = pm)
  
  py = rep(0, D)
  py = 1/(1+exp(-theta_zero_Y - theta_MY * M - theta_U1Y * U1))
  Y <- rbinom(n = D, size = 1, prob = py)
  return(Y)
}
```

```{r}
estimated_y_hmc_list_misspecify <- list()
hmc_fit_list_misspecify <- list()
means_hmc_misspecify <- c()
means_hmc_list_misspecify <- list()

start_time <- Sys.time()
for (num_data_points in ndp) {
  print("num_data_points is:")
  print(num_data_points)
  for (data_count in 1:K) {
    print("data_count is:")
    print(data_count)
    data_list <- list(D=num_data_points,
                      X1 = obs_data_list[[data_count]]$X1[1:num_data_points],
                      X2 = obs_data_list[[data_count]]$X2[1:num_data_points],
                      X3 = obs_data_list[[data_count]]$X3[1:num_data_points],
                      X4 = obs_data_list[[data_count]]$X4[1:num_data_points],
                      X5 = obs_data_list[[data_count]]$X5[1:num_data_points],
                      M = obs_data_list[[data_count]]$M[1:num_data_points],
                      Y = obs_data_list[[data_count]]$Y[1:num_data_points])
    #If you get this error (Cluster setup failed. 2 of 2 workers failed to connect.) run these lines:
    # if (Sys.getenv("RSTUDIO") == "1" && !nzchar(Sys.getenv("RSTUDIO_TERM")) &&
    #     Sys.info()["sysname"] == "Darwin") {
    #   parallel:::setDefaultClusterOptions(setup_strategy = "sequential")
    # }
    hmc_fit_list_misspecify[[data_count]] <- rstan::sampling(mod_misspecify, data=data_list, chains = 2, iter = 3000, warmup = 1500, seed = 5, control = list(max_treedepth = 15)) #seed2,5
    samples_hmc_misspecify <- rstan::extract(hmc_fit_list_misspecify[[data_count]], c("mu1", "u1_scale",
                                                                                      "theta_zero_X1", "theta_U1X1", 
                                                                                      "theta_zero_X2", "theta_U1X2", 
                                                                                      "theta_zero_X3", "theta_U1X3", 
                                                                                      "theta_zero_X4", "theta_U1X4", 
                                                                                      "theta_zero_X5", "theta_U1X5", 
                                                                                      "theta_zero_M","theta_X1M", "theta_X2M", "theta_X3M","theta_X4M","theta_X5M",
                                                                                      "theta_zero_Y", "theta_MY","theta_U1Y"))
    estimated_y_hmc_list_misspecify[[data_count]] <-  mutilated_model_misspecify(mu1 = mean(samples_hmc_misspecify$mu1),
                                                                                 u1_scale = mean(samples_hmc_misspecify$u1_scale),
                                                                                 theta_zero_M = mean(samples_hmc_misspecify$theta_zero_M),
                                                                                 theta_X1M = mean(samples_hmc_misspecify$theta_X1M),
                                                                                 theta_X2M = mean(samples_hmc_misspecify$theta_X2M),
                                                                                 theta_X3M = mean(samples_hmc_misspecify$theta_X3M),
                                                                                 theta_X4M = mean(samples_hmc_misspecify$theta_X4M),
                                                                                 theta_X5M = mean(samples_hmc_misspecify$theta_X5M),
                                                                                 theta_zero_Y = mean(samples_hmc_misspecify$theta_zero_Y),
                                                                                 theta_MY = mean(samples_hmc_misspecify$theta_MY),
                                                                                 theta_U1Y = mean(samples_hmc_misspecify$theta_U1Y),
                                                                                 x1 = x1[1:num_generated_samples],
                                                                                 x2 = x2[1:num_generated_samples],
                                                                                 x3 = x3[1:num_generated_samples],
                                                                                 x4 = x4[1:num_generated_samples],
                                                                                 x5 = x5[1:num_generated_samples],
                                                                                 seed = seed,
                                                                                 num_generated_samples = num_generated_samples
    )
    means_hmc_misspecify <- c(means_hmc_misspecify,abs(mean(estimated_y_hmc_list_misspecify[[data_count]]) - mean(intv_data_list[[data_count]]$Y)))
  }
  means_hmc_list_misspecify[[paste0("num_data_points_",num_data_points)]] <- means_hmc_misspecify
  print("means_hmc_list_misspecify is:")
  print(means_hmc_list_misspecify)
  means_hmc_misspecify <- c()
}
end_time <- Sys.time()
end_time - start_time
#saveRDS(means_hmc_list_misspecify, file = "data/Case1/output/means_hmc_list_misspecify_Case1.RData")
```

# Plug-in 

In this section we want to compare our approach with the plug-in estimator. We have to construct a parametric model for the conditional distributions that appear in the formula of an identifying functional of a causal effect. For case study 1, the do-calculus based formula is as follows:

$$
P_{M_G}(Y | do(\textbf{x}')) = \int_{m, \textbf x} P_{M_G}(Y|m, \textbf{x}) P_{M_G}(m | \textbf{x}') P_{M_G}(\textbf{x}) dm \text{ } d\textbf{x}
$$
The Stan model is as follows:

```{r}
model_str_plug_in <- "
    data {
        int D;
        int<lower=0,upper=1> X1[D];
        int<lower=0,upper=1> X2[D];
        int<lower=0,upper=1> X3[D];
        int<lower=0,upper=1> X4[D];
        int<lower=0,upper=1> X5[D];
        int<lower=0,upper=1> M[D];
        int<lower=0,upper=1> Y[D];
    }
  parameters {
       real theta_zero_X1; 
       real theta_zero_X2; 
       real theta_zero_X3; 
       real theta_zero_X4; 
       real theta_zero_X5; 
       real theta_zero_M; 
       real theta_X1M;
       real theta_X2M; 
       real theta_X3M;
       real theta_X4M;
       real theta_X5M;
       real theta_zero_Y;
       real theta_MY;
       real theta_X1Y;
       real theta_X2Y;
       real theta_X3Y;
       real theta_X4Y;
       real theta_X5Y;
    }

    model {
        //prior over parameters
        theta_zero_X1  ~ normal(0, 10); 
        theta_zero_X2  ~ normal(0, 10); 
        theta_zero_M ~ normal(0, 10); 
        theta_X1M ~ normal(0, 10); 
        theta_X2M ~ normal(0, 10); 
        theta_zero_Y ~ normal(0, 10); 
        theta_MY ~ normal(0, 10); 
        theta_X1Y ~ normal(0, 10); 
        theta_X2Y ~ normal(0, 10); 
        theta_X3Y ~ normal(0, 10);
        theta_X4Y ~ normal(0, 10);
        theta_X5Y ~ normal(0, 10);
        //likelihood
        X1 ~ bernoulli_logit(theta_zero_X1); 
        X2 ~ bernoulli_logit(theta_zero_X2); 
        X3 ~ bernoulli_logit(theta_zero_X3); 
        X4 ~ bernoulli_logit(theta_zero_X4); 
        X5 ~ bernoulli_logit(theta_zero_X5); 
        for(i in 1:D) {
            M[i] ~ bernoulli_logit(theta_zero_M + theta_X1M * X1[i] + theta_X2M * X2[i] + theta_X3M * X3[i] + theta_X4M * X4[i] + theta_X5M * X5[i]);  
            Y[i] ~ bernoulli_logit(theta_zero_Y + theta_MY * M[i] + theta_X1Y * X1[i] + theta_X2Y * X2[i] + theta_X3Y * X3[i] + theta_X4Y * X4[i] + theta_X5Y * X5[i]); 
        }
    }
"
```


```{r, message=FALSE, warning=FALSE}
mod_plug_in <- rstan::stan_model(model_code = model_str_plug_in)
```


# Mutilated model

```{r}
mutilated_model_plug_in <- function(theta_zero_M,theta_X1M,theta_X2M,theta_X3M,theta_X4M,theta_X5M,
                                    theta_zero_Y, theta_MY,
                                    theta_X1Y,theta_X2Y,theta_X3Y,theta_X4Y,theta_X5Y,
                                    x1, x2, x3, x4, x5,
                                    seed, num_generated_samples) {
  set.seed(seed)
  
  X1 <- x1
  X2 <- x2
  X3 <- x3
  X4 <- x4
  X5 <- x5
  
  pm = rep(0, D)
  pm = 1/(1+exp(-theta_zero_M - theta_X1M * X1 - theta_X2M * X2 - theta_X3M * X3 - theta_X4M * X4 - theta_X5M * X5))
  M <- rbinom(n = D, size = 1, prob = pm)
  
  py = rep(0, D)
  py = 1/(1+exp(-theta_zero_Y - theta_MY * M - theta_X1Y * X1 - theta_X2Y * X2 - theta_X3Y * X3 - theta_X4Y * X4 - theta_X5Y * X5))
  Y <- rbinom(n = D, size = 1, prob = py)
  return(Y)
}
```


```{r}
estimated_y_hmc_list_plug_in <- list()
hmc_fit_list_plug_in <- list()
means_hmc_plug_in <- c()
means_hmc_list_plug_in <- list()
start_time <- Sys.time()
for (num_data_points in ndp) {
  print("num_data_points is:")
  print(num_data_points)
  for (data_count in 1:K) {
    print("data_count is:")
    print(data_count)
    data_list <- list(D=num_data_points,
                      X1 = obs_data_list[[data_count]]$X1[1:num_data_points],
                      X2 = obs_data_list[[data_count]]$X2[1:num_data_points],
                      X3 = obs_data_list[[data_count]]$X3[1:num_data_points],
                      X4 = obs_data_list[[data_count]]$X4[1:num_data_points],
                      X5 = obs_data_list[[data_count]]$X5[1:num_data_points],
                      M = obs_data_list[[data_count]]$M[1:num_data_points],
                      Y = obs_data_list[[data_count]]$Y[1:num_data_points])
    #If you get this error (Cluster setup failed. 2 of 2 workers failed to connect.) run these lines:
    # if (Sys.getenv("RSTUDIO") == "1" && !nzchar(Sys.getenv("RSTUDIO_TERM")) &&
    #     Sys.info()["sysname"] == "Darwin") {
    #   parallel:::setDefaultClusterOptions(setup_strategy = "sequential")
    # }
    hmc_fit_list_plug_in[[data_count]] <- rstan::sampling(mod_plug_in, data=data_list, chains = 2, iter = 3000, warmup = 1500, seed = 2, control = list(max_treedepth = 15)) #seed2,5
    samples_hmc_plug_in <- rstan::extract(hmc_fit_list_plug_in[[data_count]], c("theta_zero_X1", "theta_zero_X2","theta_zero_X3","theta_zero_X4","theta_zero_X5",
                                                                                "theta_zero_M","theta_X1M", "theta_X2M","theta_X3M","theta_X4M","theta_X5M",
                                                                                "theta_zero_Y", "theta_MY","theta_X1Y","theta_X2Y","theta_X3Y","theta_X4Y","theta_X5Y"))
    estimated_y_hmc_list_plug_in[[data_count]] <-  mutilated_model_plug_in(
                                            theta_zero_M = mean(samples_hmc_plug_in$theta_zero_M),
                                            theta_X1M = mean(samples_hmc_plug_in$theta_X1M),
                                            theta_X2M = mean(samples_hmc_plug_in$theta_X2M),
                                            theta_X3M = mean(samples_hmc_plug_in$theta_X3M),
                                            theta_X4M = mean(samples_hmc_plug_in$theta_X4M),
                                            theta_X5M = mean(samples_hmc_plug_in$theta_X5M),
                                            theta_zero_Y = mean(samples_hmc_plug_in$theta_zero_Y),
                                            theta_MY = mean(samples_hmc_plug_in$theta_MY),
                                            theta_X1Y = mean(samples_hmc_plug_in$theta_X1Y),
                                            theta_X2Y = mean(samples_hmc_plug_in$theta_X2Y),
                                            theta_X3Y = mean(samples_hmc_plug_in$theta_X3Y),
                                            theta_X4Y = mean(samples_hmc_plug_in$theta_X4Y),
                                            theta_X5Y = mean(samples_hmc_plug_in$theta_X5Y),
                                            x1 = x1[1:num_generated_samples],
                                            x2 = x2[1:num_generated_samples],
                                            x3 = x3[1:num_generated_samples],
                                            x4 = x4[1:num_generated_samples],
                                            x5 = x5[1:num_generated_samples],
                                            seed = seed,
                                            num_generated_samples = num_generated_samples
    )
    means_hmc_plug_in <- c(means_hmc_plug_in,abs(mean(estimated_y_hmc_list_plug_in[[data_count]]) - mean(intv_data_list[[data_count]]$Y)))
    
  }
  means_hmc_list_plug_in[[paste0("num_data_points_",num_data_points)]] <- means_hmc_plug_in
  print("means_hmc_list_plug_in is:")
  print(means_hmc_list_plug_in)
  means_hmc_plug_in <- c()
}
end_time <- Sys.time()
end_time - start_time
#saveRDS(means_hmc_list_plug_in, file = "data/Case1/output/means_hmc_list_plug_in_Case1.RData")
```