---
title: "GRN of E.Coli: Conditional Independencies vs Sample Size"
output: html_document
date: "2023-03-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages

```{r, echo=FALSE}
library(dagitty)
library(lavaan)
library(sna)
library(DOT)
library(igraph)
library(pcalg)
library(SEMgraph)
library(reticulate)
library(dplyr)
```

## Load E.coli GRN graph and data

The data are from SERGIO project. The file contain 100 data for 100 genes with 300 cells each with 3 cell types (i.e., conditions)

```{r}
dot <- read.dot("~/Missing data/SERGIO/GNW_sampled_GRNs/Ecoli_100_net1.dot") # GRN graph
# GRN_out <- read.csv("~/Missing data/SERGIO/Demo/GeneExpression_SingleCell.csv") # Simulated single cell data
GRN_out <- read.csv("~/Missing data/SERGIO/data_sets/De-noised_100G_3T_300cPerT_dynamics_8_DS8/simulated_noNoise_S_0.csv") # Existing data
GeneIDs <- read.delim("~/Missing data/GRN_Analysis/GeneIDs.txt", header=FALSE) # Gene IDs
GeneIDs <- gsub(';','',GeneIDs[,2])
#summary(GRN_out)
GRN_out <- t(GRN_out[,-1])
colnames(GRN_out) <- GeneIDs
dim(GRN_out)
head(GRN_out)
```

# Pearson's correlation plot (Linear correlation)

```{r}
res <- cor(GRN_out)
corrplot::corrplot(res,type = "upper", tl.col = "black", tl.cex = 0.5)
```

# Create data frames for data by cell type

```{r}
df <- data.frame(GRN_out) # 100 genes with 300 cells with 3 types
df.celltype1 <- df %>%
  slice(which(row_number() %% 3 == 1)) # Cell type 1
df.celltype2 <- df %>%
  slice(which(row_number() %% 3 == 2)) # Cell type 2
df.celltype3 <- df %>%
  slice(which(row_number() %% 3 == 0)) # Cell type 3


```

# The GRN graph
```{r}
g.graph <- graph.adjacency(dot)
plot(g.graph)
g <- graph2dagitty(g.graph)
g <- graphLayout(g)
plot(g)
```


## Conditional independencies

```{r}
# For each pair of non-adjacent nodes in this graph, the set of variables that d-separates that pair.
# i.e., for each non-adjacent pair of variables, all minimal sets that we can condition on to render that pair independent.
impliedConditionalIndependencies(g, type = "missing.edge", max.results = Inf)

# Draw paths from icd to fadI
paths(g, "icd", "fadI" )$paths

# Draw directed paths from icd to fadI
paths(g, "potI", "nemA", directed=TRUE)$paths


# Conditional independencies from the GRN
true.ind <- impliedConditionalIndependencies(g, type = "missing.edge", max.results = Inf)

```

## Non-linear correlations

From all the existing conditional independencies in the GRN, the proportion of independencies that are shown as conditionally dependent from data is calculated here.

### Cell type 1

```{r}
# Compute independencies with at most 3 conditioning variables
d <- df.celltype1
imp <- Filter(function(x) length(x$Z)<4, impliedConditionalIndependencies(g))
CI <- localTests( g, d, "cis.loess", R=100, tests=imp, loess.pars=list(span=0.6) )
plotLocalTestResults (head(CI,50))

ind <- CI[(CI$`2.5%` <0 & CI$`97.5%`>0),]
dep <- CI[!(CI$`2.5%` <0 & CI$`97.5%`>0),]

# Plot the conditional independencies that we were not able to prove to be dependent (we cannot conclude them to be independent: do not reject null hypothesis that beta=0)
plotLocalTestResults (ind)

# Plot the conditional independencies that we prove to be dependent
plotLocalTestResults (dep)

# When sample size increases the probability of detecting dependencies increases. 
p <- nrow(dep)/(nrow(dep)+nrow(ind))
p

```

### Cell type 2

```{r}
# Compute independencies with at most 3 conditioning variables
d <- df.celltype2
imp <- Filter(function(x) length(x$Z)<4, impliedConditionalIndependencies(g))
CI <- localTests( g, d, "cis.loess", R=100, tests=imp, loess.pars=list(span=0.6) )
plotLocalTestResults (head(CI,50))

ind <- CI[(CI$`2.5%` <0 & CI$`97.5%`>0),]
dep <- CI[!(CI$`2.5%` <0 & CI$`97.5%`>0),]

# Plot the conditional independencies that we were not able to prove to be dependent (we cannot conclude them to be independent: do not reject null hypothesis that beta=0)
plotLocalTestResults (ind)

# Plot the conditional independencies that we prove to be dependent
plotLocalTestResults (dep)

# When sample size increases the probability of detecting dependencies increases. 
p <- nrow(dep)/(nrow(dep)+nrow(ind))
p

```

### Cell type 3

```{r}
# Compute independencies with at most 3 conditioning variables
d <- df.celltype3
imp <- Filter(function(x) length(x$Z)<4, impliedConditionalIndependencies(g))
CI <- localTests( g, d, "cis.loess", R=100, tests=imp, loess.pars=list(span=0.6) )
plotLocalTestResults (head(CI,50))

ind <- CI[(CI$`2.5%` <0 & CI$`97.5%`>0),]
dep <- CI[!(CI$`2.5%` <0 & CI$`97.5%`>0),]

# Plot the conditional independencies that we were not able to prove to be dependent (we cannot conclude them to be independent: do not reject null hypothesis that beta=0)
plotLocalTestResults (ind)

# Plot the conditional independencies that we prove to be dependent
plotLocalTestResults (dep)

# When sample size increases the probability of detecting dependencies increases. 
p <- nrow(dep)/(nrow(dep)+nrow(ind))
p


```


